---
title: "POLISCI450D Week1 Practice"
subtitle: "Modern Difference-in-Differences"
author: "Rabia Kutlu Karasu"
format:
  html:
    toc: true
    code-fold: false
    embed-resources: true
jupyter: r
---

## Introduction

We will replicate two papers using the **fect** package for counterfactual estimation in panel data today:

1. **Hainmueller and Hangartner (2019)**: Examining the effects of indirect versus direct democracy on naturalization rates in Switzerland (staggered adoption design **without treatment reversals**)
2. **Grumbach and Sahn (2020)**: Investigating the mobilizing effect of minority candidates on coethnic support in U.S. congressional elections (design **with treatment reversals**)

For additional practice, Yan (2025) (on the syllabus) offers another useful application. I have the replication files and am happy to share them with anyone who would like to practice further and sharpen their skills.

The **fect** package implements counterfactual (imputation) estimators for causal inference with time-series cross-sectional data. These methods address limitations of traditional two-way fixed effects (TWFE) models when dealing with heterogeneous treatment effects and staggered adoption.


### Installing Required Packages

First, install the necessary packages from CRAN and GitHub:


```{r install-packages, eval=FALSE}
rm(list = ls())
devtools::install_github("xuyiqing/fect")
installed.packages()["fect", "Version"]
devtools::install_github('xuyiqing/panelView')
install_all <- function(packages) {
  installed_pkgs <- installed.packages()[, "Package"]
  for (pkg in packages) {
    if (!pkg %in% installed_pkgs) {
      install.packages(pkg)
    }
  }
}
if (!"HonestDiDFEct" %in% rownames(installed.packages())) {
  devtools::install_github("lzy318/HonestDiDFEct")
}

packages <- c("abind", "doParallel", "doRNG", "fixest", "foreach", "future", 
              "GGally", "ggplot2", "grid", "gridExtra", "Mass", 
              "panelView", "Rcpp")
install_all(packages)

packages <- c("dplyr", "fixest", "did", "didimputation", 
              "panelView", "ggplot2", "bacondecomp", "HonestDiD",
              "DIDmultiplegtDYN", "PanelMatch","patchwork")
install.packages(setdiff(packages, rownames(installed.packages())))  

```

### Loading Libraries

```{r load-libraries, message=FALSE, warning=FALSE}

library(dplyr)
library(readstata13)
library(fixest)
library(did)
library(fect)
library(panelView)
library(PanelMatch)
library(ggplot2)
library(bacondecomp)
library(fect)
library(didimputation)
library(doParallel)
library(HonestDiD)
library(HonestDiDFEct)
library(patchwork)

```

### Loading Datasets

The **fect** package includes built-in datasets for both examples:

```{r load-data}
data(fect)
ls()  # Lists available datasets

# View the first few rows of each dataset
head(hh2019) #Hainmueller and Hangartner (2019) data
head(gs2020) #Grumbach and Sahn (2020) data
```

## Example 1: Hainmueller and Hangartner (2019)

### Research Question

Does switching from direct democracy to indirect democracy increase naturalization rates for minority immigrants in Swiss municipalities?

### Data Structure

The `hh2019` dataset contains:
- **bfs**: Municipality identifier
- **year**: Time period (1991-2009)
- **nat_rate_ord**: Naturalization rate (outcome variable)
- **indirect**: Treatment indicator (1 = indirect democracy, 0 = direct democracy)


### Visualizing Treatment Status

This is a staggered adoption design where different municipalities switch from direct to indirect democracy at different times, with no treatment reversals.

```{r hh-treatment-viz, fig.width=8, fig.height=6}
panelview(nat_rate_ord ~ indirect, 
          data = hh2019, 
          index = c("bfs", "year"),
          axis.lab = "time",
          xlab = "Year", 
          ylab = "Municipality",
          gridOff = TRUE,
          by.timing = TRUE,
          background = "white",
          main = "Hainmueller & Hangartner (2019): Treatment Status")
```

### Visualizing Outcome Variable

```{r hh-outcome-viz, fig.width=8, fig.height=6}
panelview(nat_rate_ord ~ indirect, 
          data = hh2019, 
          index = c("bfs", "year"),
          type = "outcome",
          xlab = "Year", 
          ylab = "Municipality",
          main = "Hainmueller & Hangartner (2019): Naturalization Rates")
```

### Estimating Treatment Effects

We use the Fixed Effects Counterfactual estimator (FEct), which is equivalent to the "imputation method" or "two-stage DID":

```{r hh-fect-estimation}
# Estimate with FEct method
out.hh <- fect(nat_rate_ord ~ indirect, 
               data = hh2019, 
               index = c("bfs", "year"),
               method = "fe",           # Fixed effects counterfactual
               force = "two-way",       # Two-way fixed effects
               se = TRUE,               # Calculate standard errors
               vartype = "bootstrap",   # Bootstrap inference
               nboots = 200,            # Number of bootstrap iterations
               parallel = TRUE)         # Use parallel computing

# View summary
print(out.hh)
```

### Visualizing Results: Gap Plot

The gap plot shows estimated dynamic treatment effects (ATT) over time:

```{r hh-gap-plot, fig.width=10, fig.height=6}
plot(out.hh, 
     type = "gap",
     main = "Hainmueller & Hangartner (2019): Dynamic Treatment Effects",
     ylab = "Effect on Naturalization Rate",
     xlab = "Time Relative to Treatment")
```

### Visualizing Results: Counterfactual Plot

The counterfactual plot shows the observed outcomes for treated units versus the predicted counterfactual (what would have happened without treatment):

```{r hh-counterfactual-plot, fig.width=10, fig.height=6}
plot(out.hh, 
     type = "counterfactual",
     main = "Hainmueller & Hangartner (2019): Treated vs. Counterfactuals",
     ylab = "Naturalization Rate",
     legend.pos = "bottom")
```

### Diagnostic Tests

Conduct placebo tests to validate the parallel trends assumption:

```{r hh-placebo-test}
# Run placebo test using 3 pre-treatment periods
out.hh.placebo <- fect(nat_rate_ord ~ indirect, 
                       data = hh2019, 
                       index = c("bfs", "year"),
                       method = "fe",
                       force = "two-way",
                       se = TRUE,
                       vartype = "bootstrap",
                       nboots = 200,
                       placeboTest = TRUE,
                       placebo.period = c(-3, -1))  # Use periods -3 to -1 as placebo

# Visualize placebo test results
plot(out.hh.placebo, 
     type = "gap",
     main = "Placebo Test: Hainmueller & Hangartner (2019)")
```

### Interpretation

Based on the analysis:

1. **Main Finding**: Switching from direct to indirect democracy increased naturalization rates by approximately 1.2-1.7 percentage points on average
2. **Dynamic Effects**: The treatment effect persists and may strengthen over time
3. **Parallel Trends**: Placebo tests should show no significant pre-treatment effects, supporting the identifying assumptions



## Example 2: Grumbach and Sahn (2020)

### Research Question

Does the presence of an Asian candidate mobilize coethnic support (measured by the proportion of campaign contributions from Asian donors) in U.S. congressional elections?

### Data Structure

The `gs2020` dataset contains:
- **district**: Congressional district identifier
- **cycle**: Election cycle (time period)
- **pct_asian_gen**: Proportion of contributions from Asian donors (outcome)
- **asian_cand**: Presence of Asian candidate (treatment indicator)

This design includes **treatment reversals**, where districts can have Asian candidates in some elections but not others.

### Visualizing Treatment Status

```{r gs-treatment-viz, fig.width=8, fig.height=6}

panelview(Y="general_sharetotal_A_all", D="cand_A_all",
          X=c("cand_H_all", "cand_B_all"), 
          index = c("district_final", "cycle"),
          data = gs2020, 
          xlab = "Time Period", ylab = "District",
          gridOff = TRUE, 
          by.timing = TRUE,
          cex.legend=5, cex.axis= 5, 
          cex.main = 10, cex.lab = 5,
          main = "Grumbach & Sahn (2020): Treatment Status (Asian Candidate Presence)")
```

### Visualizing Outcome Variable

```{r gs-outcome-viz, fig.width=8, fig.height=6}
panelview(general_sharetotal_A_all ~ cand_A_all, 
          data = gs2020, 
          index = c("district_final", "cycle"),
          type = "outcome",
          xlab = "Election Cycle", 
          ylab = "District",
          main = "Grumbach & Sahn (2020): Proportion of Asian Donations")
```

### Estimating Treatment Effects with Treatment Reversals

The FEct method can accommodate treatment reversals under the assumption of limited carryover effects:

```{r gs-fect-estimation}
# Estimate with FEct method
out.gs <- fect(general_sharetotal_A_all ~ cand_A_all, 
               data = gs2020, 
               index = c("district_final", "cycle"),
               method = "fe",
               force = "two-way",
               se = TRUE,
               vartype = "bootstrap",
               nboots = 200,
               parallel = TRUE)

# View summary
print(out.gs)
```

### Visualizing Results: Gap Plot

```{r gs-gap-plot, fig.width=10, fig.height=6}
plot(out.gs, 
     type = "gap",
     main = "Grumbach & Sahn (2020): Dynamic Treatment Effects",
     ylab = "Effect on Asian Donor Proportion",
     xlab = "Time Relative to Treatment")
```

### Visualizing Results: Counterfactual Plot

```{r gs-counterfactual-plot, fig.width=10, fig.height=6}
plot(out.gs, 
     type = "counterfactual",
     main = "Grumbach & Sahn (2020): Treated vs. Counterfactuals",
     ylab = "Proportion of Asian Donations",
     legend.pos = "bottom")
```

### Testing for Carryover Effects

When treatment reverses, we should test whether effects persist after treatment ends:

```{r gs-carryover-test}
# Run carryover test
out.gs.carryover <- fect(general_sharetotal_A_all ~ cand_A_all, 
                         data = gs2020, 
                         index = c("district_final", "cycle"),
                         method = "fe",
                         force = "two-way",
                         se = TRUE,
                         vartype = "bootstrap",
                         nboots = 200,
                         carryoverTest = TRUE,
                         carryover.period = c(1, 3))  # Test 1-3 periods after treatment ends

# Visualize carryover test
plot(out.gs.carryover, 
     type = "exit",
     main = "Carryover Effects Test: Grumbach & Sahn (2020)")
```
So, limited evidence of carryover effects suggests the treatment effect is contemporaneous.


## Sensitivity Analysis

Following Rambachan and Roth (2023), we can assess sensitivity to potential parallel trends violations:

```{r sensitivity-analysis}
# Run fect with placebo periods
out.fect.placebo <- fect(nat_rate_ord~indirect, data = hh2019, 
                         index = c("bfs","year"),
                         method = 'fe', se = TRUE, 
                         placeboTest = TRUE, placebo.period = c(-2,0))

# Define post-treatment periods and sensitivity parameters for fect_sens
T.post <- 10 # Number of post-treatment periods based on original analysis
post_periods_vec <- 1:T.post

# Parameters for Relative Magnitude (RM) restriction
Mbar_vec_avg_rm <- seq(0, 1, by = 0.1)    # For average ATT plot
Mbar_vec_period_rm <- c(0, 0.5)          # For period-by-period ATT plot

# Parameters for Smoothness restriction
M_vec_avg_smooth <- seq(0, 0.25, by = 0.05) # For average ATT plot
M_vec_period_smooth <- c(0, 0.1)           # For period-by-period ATT plot

# Run sensitivity analysis using fect_sens
# This function augments out.fect.placebo with sensitivity results
out.fect.placebo <- fect_sens(
  fect.out      = out.fect.placebo,
  post.periods  = post_periods_vec,
  Mbarvec       = Mbar_vec_avg_rm,
  periodMbarvec = Mbar_vec_period_rm,
  Mvec          = M_vec_avg_smooth,
  periodMvec    = M_vec_period_smooth,
  parallel      = TRUE # Set to TRUE for parallel processing if desired
)

plot(out.fect.placebo,
     type = "sens",
     restrict = "rm",
     main = "Sensitivity Analysis: Relative Magnitude Restriction")

```

## Comparing with Other Modern DID Estimators

The **fect** package provides a wrapper function to compare multiple estimators:

```{r did-wrapper, warning=FALSE, message=FALSE}

# Compare multiple estimators
res_twfe_es <- did_wrapper(
  data   = hh2019,
  Y      = "nat_rate_ord",
  D      = "indirect",
  index  = c("bfs", "year"),
  method = "twfe",
  se     = "default"
)

res_stacked_es <- did_wrapper(
  data   = hh2019,
  Y      = "nat_rate_ord",
  D      = "indirect",
  index  = c("bfs", "year"),
  method = "st",
  se     = "default"
)

res_stacked_es_b <- did_wrapper(
  data   = hh2019,
  Y      = "nat_rate_ord",
  D      = "indirect",
  index  = c("bfs", "year"),
  method = "st",
  se     = "boot"
)

res_iw_es <- did_wrapper(
  data   = hh2019,
  Y      = "nat_rate_ord",
  D      = "indirect",
  index  = c("bfs", "year"),
  method = "iw",
  se     = "default"
)

p_twfe<- esplot(res_twfe_es, main="",xlab="")
p_stacked<- esplot(res_stacked_es, main="",xlab="")
p_stacked_b <- esplot(res_stacked_es_b, main="",xlab="")
p_iw<- esplot(res_iw_es, main="",xlab="")


(p_twfe | p_stacked | p_iw) +
  plot_annotation(
    title = "Event Study Comparison Across DID Estimators",
    subtitle = "Same data, same event time — different estimators",
    caption = "TWFE, Stacked DID, and Interaction-Weighted estimators"
  )

(p_stacked_b|p_stacked)+
  plot_annotation(
    title = "Stacked DID",
    caption = "Stacked DID with and without bootstrapped standard errors."
  )
```


## References

- Chiu, A., Lan, X., Liu, Z., & Xu, Y. (2025). Causal Panel Analysis Under Parallel Trends: Lessons from a Large Reanalysis Study. *American Political Science Review*.

- Grumbach, J. M., & Sahn, A. (2020). Race and Representation in Campaign Finance. *American Political Science Review*, 114(1), 206-221.

- Hainmueller, J., & Hangartner, D. (2019). Does Direct Democracy Hurt Immigrant Minorities? Evidence from Naturalization Decisions in Switzerland. *American Journal of Political Science*, 63(3), 530-547.

- Liu, L., Wang, Y., & Xu, Y. (2024). A Practical Guide to Counterfactual Estimators for Causal Inference with Time-Series Cross-Sectional Data. *American Journal of Political Science*, 68(1), 160-176.

- Rambachan, A., & Roth, J. (2023). A More Credible Approach to Parallel Trends. *Review of Economic Studies*, rdad018.

- Xu, Y. (2017). Generalized Synthetic Control Method: Causal Inference with Interactive Fixed Effects Models. *Political Analysis*, 25(1), 57-76.

## Additional Resources

- **fect package documentation**: https://yiqingxu.org/packages/fect
- **GitHub repository**: https://github.com/xuyiqing/fect
- **panelView package**: https://github.com/xuyiqing/panelView
