---
title: "Synthetic Control Methods: The Economic Costs of Conflict"
subtitle: "Replication and Modern Extensions"
author: "Rabia Kutlu Karasu"
format:
  html:
    toc: true
    embed-resources: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
    highlight-style: github
    type: book
    output-dir: docs
execute:
  warning: false
  message: false
website:
  site-url: "https://github.com/rabiakutlukarasu/Polisci450D_2026winter.git"
---

# Introduction

This chapter replicates the seminal synthetic control analysis from **Abadie and Gardeazabal (2003)** examining the economic impact of terrorism in the Basque Country, then extends the analysis using modern methods:

- **Generalized Synthetic Control (GSynth)** (Xu 2017) - Interactive fixed effects model
- **Synthetic Difference-in-Differences** (Arkhangelsky et al. 2021)
- **Augmented Synthetic Control** (Ben-Michael, Feller, and Rothstein 2021)  
- **Triply Robust Panel Estimators** (Athey et al. 2025)

## The Research Question

During the late 1960s and 1970s, the Basque terrorist organization ETA (Euskadi Ta Askatasuna) intensified its violent campaign for independence from Spain. **Did this conflict reduce economic growth in the Basque Country?**

## Why Synthetic Control?

Traditional difference-in-differences requires parallel trends - but what if no single control region provides a good comparison? The Basque Country has unique characteristics:

- Industrial economy
- Cultural distinctiveness  
- Geographic location
- Political history

**Solution**: Create a "synthetic" Basque Country as a weighted combination of other Spanish regions that resembles pre-terrorism Basque Country on economic indicators.


## Setup and Data


```{r setup}
# Load required packages
library(Synth)           # Original synthetic control
library(tidyverse)      
library(fect)            # Generalized synthetic control (Xu 2017)
library(synthdid)        # Synthetic DiD (Arkhangelsky et al. 2021)
library(augsynth)        # Augmented synthetic control (Ben-Michael et al. 2021)
library(knitr)        
library(kableExtra)     
library(patchwork)  
library(panelView)

# Set theme for all plots
theme_set(theme_bw(base_size = 12) +
          theme(plot.title = element_text(face = "bold", size = 14),
                plot.subtitle = element_text(size = 11),
                legend.position = "bottom"))
```


```{r load-data}
# Load the Basque Country data (built into Synth package)
data("basque")

# Examine the data structure
glimpse(basque)

# Create a summary table of the data
basque %>%
  group_by(regionname) %>%
  summarise(
    years = paste(min(year), "-", max(year)),
    n_obs = n(),
    avg_gdp = round(mean(gdpcap, na.rm = TRUE), 0)
  ) %>%
  head(10) %>%
  kable(
    col.names = c("Region", "Years", "Observations", "Avg GDP per Capita"),
    caption = "Sample of Spanish Regional Panel Data"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))
```

The dataset includes:
- **17 Spanish regions** (excluding aggregate Spain)
- **Time period**: 1955-1997
- **Variables**: 
  - GDP per capita (outcome)
  - Education levels (predictors)
  - Sector composition (predictors)
  - Investment rates (predictor)
  
  
## Visualizing the Problem

```{r viz-raw-trends, fig.width=10, fig.height=6}
# Plot GDP trends for Basque Country vs other regions
basque %>%
  mutate(
    region_type = case_when(
      regionname == "Basque Country (Pais Vasco)" ~ "Basque Country",
      regionname == "Spain (Espana)" ~ "Spain (aggregate)",
      TRUE ~ "Other regions"
    )
  ) %>%
  ggplot(aes(x = year, y = gdpcap, group = regionname, color = region_type)) +
  geom_line(alpha = 0.6, linewidth = 0.8) +
  geom_vline(xintercept = 1975, linetype = "dashed", color = "red", linewidth = 1) +
  annotate("text", x = 1975, y = max(basque$gdpcap, na.rm = TRUE), 
           label = "Death of Franco (1975)\nIntensification of terrorism", 
           hjust = -0.05, vjust = 1, size = 3.5) +
  scale_color_manual(values = c("Basque Country" = "#D55E00", 
                                 "Spain (aggregate)" = "#0072B2",
                                 "Other regions" = "gray70"),
                     name = "") +
  labs(
    title = "GDP per Capita: Basque Country vs Spanish Regions",
    subtitle = "Raw data shows divergence after 1975, but which regions are comparable?",
    x = "Year",
    y = "GDP per Capita (thousands of pesetas)",
    caption = "Source: Abadie & Gardeazabal (2003)"
  ) +
  theme(legend.position = "top")
```

After 1975, the Basque Country's GDP per capita trajectory appears to diverge from other regions.


# Part 1: Classic Synthetic Control Method

## Preparing the Data

The `Synth` package requires data in a specific format using the `dataprep()` function:

```{r dataprep}
# Prepare data for Synth package
dataprep.out <- dataprep(
  foo = basque,
  predictors = c("school.illit", "school.prim", "school.med",
                 "school.high", "school.post.high", "invest"),
  predictors.op = "mean",
  time.predictors.prior = 1964:1969,
  special.predictors = list(
    list("gdpcap", 1960:1969, "mean"),
    list("sec.agriculture", seq(1961, 1969, 2), "mean"),
    list("sec.energy", seq(1961, 1969, 2), "mean"),
    list("sec.industry", seq(1961, 1969, 2), "mean"),
    list("sec.construction", seq(1961, 1969, 2), "mean"),
    list("sec.services.venta", seq(1961, 1969, 2), "mean"),
    list("sec.services.nonventa", seq(1961, 1969, 2), "mean"),
    list("popdens", 1969, "mean")
  ),
  dependent = "gdpcap",
  unit.variable = "regionno",
  unit.names.variable = "regionname",
  time.variable = "year",
  treatment.identifier = 17,  # Basque Country
  controls.identifier = c(2:16, 18),  # All other regions except Spain aggregate
  time.optimize.ssr = 1960:1969,
  time.plot = 1955:1997
)

# Run synthetic control
synth.out <- synth(data.prep.obj = dataprep.out, method = "BFGS")

# Get the results
synth.tables <- synth.tab(dataprep.res = dataprep.out, synth.res = synth.out)
```

## Synthetic Control Weights

Which regions form the "synthetic" Basque Country?

```{r synth-weights}
# Extract and display weights
weights_df <- data.frame(
  Region = rownames(synth.out$solution.w),
  Weight = round(as.numeric(synth.out$solution.w), 4)
) %>%
  dplyr::filter(Weight > 0.001) %>%
  dplyr::arrange(dplyr::desc(Weight))

weights_df

```

The synthetic Basque Country is primarily composed of Cataluna and Madrid. These weights are chosen to best match pre-treatment characteristics.

## Balance Table

How well does the synthetic control match the Basque Country before treatment?

```{r balance-table}
# Create balance table
balance_table <- data.frame(
  Variable = rownames(synth.tables$tab.pred),
  Basque = round(synth.tables$tab.pred[, 1], 2),
  Synthetic = round(synth.tables$tab.pred[, 2], 2),
  Sample = round(synth.tables$tab.pred[, 3], 2)
) %>%
  mutate(Difference = Basque - Synthetic)

# donor pool identifiers
controls <- dataprep.out$controls.identifier

# compute SDs for predictors in donor pool
pred_sds <- sapply(rownames(synth.tables$tab.pred), function(v) {
  sd(dataprep.out$X0[v, ], na.rm = TRUE)
})

balance_table <- data.frame(
  Variable = rownames(synth.tables$tab.pred),
  Basque = round(synth.tables$tab.pred[, 1], 2),
  Synthetic = round(synth.tables$tab.pred[, 2], 2),
  Sample = round(synth.tables$tab.pred[, 3], 2)
) %>%
  mutate(
    Difference = Basque - Synthetic,
    SMD = Difference / pred_sds #Standardized Mean Difference
  )

balance_table
```

The synthetic control matches the Basque Country on pre-treatment covariates except education variables.


## The Main Result: Treatment Effect

```{r synth-plot, fig.width=10, fig.height=7}
# Create the synthetic control plot
path.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "Real GDP per Capita (thousand pesetas)",
  Xlab = "Year",
  Legend = c("Basque Country", "Synthetic Basque Country"),
  Legend.position = "topleft",
  Main = "GDP per Capita: Basque Country vs Synthetic Control"
)

# Add treatment line
abline(v = 1970, lty = 2, col = "red", lwd = 2)
text(1970, max(dataprep.out$Y1plot), "Treatment begins", pos = 4, col = "red")
```

```{r gaps-plot, fig.width=10, fig.height=7}
# Plot the gap (treatment effect over time)
gaps.plot(
  synth.res = synth.out,
  dataprep.res = dataprep.out,
  Ylab = "Gap in GDP per Capita",
  Xlab = "Year",
  Main = "Treatment Effect: Gap Between Basque Country and Synthetic Control"
)

abline(h = 0, lty = 2, col = "gray50")
abline(v = 1970, lty = 2, col = "red", lwd = 2)
```


## Quantifying the Effect

```{r calculate-effects}
# Calculate average treatment effect post-1970
Y1 <- dataprep.out$Y1plot
Y0 <- dataprep.out$Y0plot %*% synth.out$solution.w

# Create a data frame of gaps
gaps_df <- data.frame(
  year = as.numeric(rownames(Y1)),
  treated = as.numeric(Y1),
  synthetic = as.numeric(Y0),
  gap = as.numeric(Y1 - Y0)
)

# Calculate ATT for different periods
att_estimates <- gaps_df %>%
  filter(year >= 1970) %>%
  summarise(
    `1970-1979` = mean(gap[year >= 1970 & year <= 1979]),
    `1980-1989` = mean(gap[year >= 1980 & year <= 1989]),
    `1990-1997` = mean(gap[year >= 1990 & year <= 1997]),
    `Overall (1970-1997)` = mean(gap)
  ) %>%
  pivot_longer(everything(), names_to = "Period", values_to = "ATT")

att_estimates %>%
  kable(
    col.names = c("Period", "Average Treatment Effect (GDP per capita)"),
    caption = "Estimated Economic Cost of Terrorism",
    digits = 4
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

On average, the Basque Country's GDP per capita was approximately **`r round(att_estimates$ATT[4], 0)` thousand pesetas lower** than it would have been without terrorism (1970-1997 average).


## Placebo Tests: In-Space Permutation

To assess whether this effect is unusual, we run placebo tests by applying the same procedure to control regions.

```{r placebo-space1, results='hide'}
# Store all control unit numbers
control_units <- c(2:16, 18)

# Initialize storage for placebo gaps
placebo_gaps <- matrix(NA, nrow = length(1955:1997), ncol = length(control_units))
rownames(placebo_gaps) <- 1955:1997

# Run synthetic control for each control unit
for(i in 1:length(control_units)) {
  
  # Prepare data with this control unit as "treated"
  tryCatch({
    dataprep_placebo <- dataprep(
      foo = basque,
      predictors = c("school.illit", "school.prim", "school.med",
                     "school.high", "school.post.high", "invest"),
      predictors.op = "mean",
      time.predictors.prior = 1964:1969,
      special.predictors = list(
        list("gdpcap", 1960:1969, "mean"),
        list("sec.agriculture", seq(1961, 1969, 2), "mean"),
        list("sec.energy", seq(1961, 1969, 2), "mean"),
        list("sec.industry", seq(1961, 1969, 2), "mean"),
        list("sec.construction", seq(1961, 1969, 2), "mean"),
        list("sec.services.venta", seq(1961, 1969, 2), "mean"),
        list("sec.services.nonventa", seq(1961, 1969, 2), "mean"),
        list("popdens", 1969, "mean")
      ),
      dependent = "gdpcap",
      unit.variable = "regionno",
      unit.names.variable = "regionname",
      time.variable = "year",
      treatment.identifier = control_units[i],
      controls.identifier = c(control_units[-i], 17),
      time.optimize.ssr = 1960:1969,
      time.plot = 1955:1997
    )
    
    synth_placebo <- synth(data.prep.obj = dataprep_placebo, method = "BFGS")
    
    # Store gaps
    Y1_placebo <- dataprep_placebo$Y1plot
    Y0_placebo <- dataprep_placebo$Y0plot %*% synth_placebo$solution.w
    placebo_gaps[, i] <- Y1_placebo - Y0_placebo
    
  }, error = function(e) {
    # Skip if synthetic control fails for this unit
    NULL
  })
}
```

```{r placebo-space2, fig.width=10, fig.height=8}
# Plot all placebo gaps
placebo_df <- as.data.frame(placebo_gaps) %>%
  mutate(year = as.numeric(rownames(.))) %>%
  pivot_longer(-year, names_to = "unit", values_to = "gap")

# Create plot
ggplot() +
  geom_line(data = placebo_df, 
            aes(x = year, y = gap, group = unit), 
            alpha = 0.3, color = "gray50") +
  geom_line(data = gaps_df, 
            aes(x = year, y = gap), 
            color = "#D55E00", linewidth = 1.5) +
  geom_vline(xintercept = 1970, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black", alpha = 0.3) +
  labs(
    title = "In-Space Placebo Test",
    subtitle = "Basque Country (orange) vs placebo gaps for all control regions (gray)",
    x = "Year",
    y = "Gap in GDP per Capita",
    caption = "Each gray line represents a 'placebo treatment' applied to a control region"
  ) +
  annotate("text", x = 1970, y = max(placebo_df$gap, na.rm = TRUE), 
           label = "Treatment", hjust = -0.1, color = "red")
```

**Interpretation**: The Basque Country's gap (orange) is much larger and more persistent than most placebo gaps (gray), suggesting the effect is unlikely to be due to chance.

## Placebo Tests: In-Time Permutation

Test whether the effect would appear if we pretend treatment occurred earlier.

```{r placebo-time1, results='hide'}
# Try treatment in 1965 instead of 1970
dataprep_time_placebo <- dataprep(
  foo = basque,
  predictors = c("school.illit", "school.prim", "school.med",
                 "school.high", "school.post.high", "invest"),
  predictors.op = "mean",
  time.predictors.prior = 1960:1964,  # Pre-treatment period for fake treatment
  special.predictors = list(
    list("gdpcap", 1960:1964, "mean"),
    list("sec.agriculture", 1960:1964, "mean"),
    list("sec.energy", 1960:1964, "mean"),
    list("sec.industry", 1960:1964, "mean"),
    list("sec.construction", 1960:1964, "mean"),
    list("sec.services.venta", 1960:1964, "mean"),
    list("sec.services.nonventa", 1960:1964, "mean"),
    list("popdens",1960:1969, "mean")
  ),
  dependent = "gdpcap",
  unit.variable = "regionno",
  unit.names.variable = "regionname",
  time.variable = "year",
  treatment.identifier = 17,
  controls.identifier = c(2:16, 18),
  time.optimize.ssr = 1960:1964,
  time.plot = 1955:1969  # Only look at pre-actual-treatment period
)

synth_time_placebo <- synth(data.prep.obj = dataprep_time_placebo, method = "BFGS")
```


```{r placebo-time2, fig.width=10, fig.height=7}
# Plot
path.plot(
  synth.res = synth_time_placebo,
  dataprep.res = dataprep_time_placebo,
  Ylab = "Real GDP per Capita",
  Xlab = "Year",
  Legend = c("Basque Country", "Synthetic (placebo)"),
  Main = "In-Time Placebo Test: Pretend Treatment in 1965"
)
abline(v = 1965, lty = 2, col = "red", lwd = 2)
```

**Interpretation**: No systematic gap appears before the actual treatment, supporting the validity of the synthetic control approach.




# Part 2: Generalized Synthetic Control (GSynth)

The **Generalized Synthetic Control** method (Xu 2017) extends the traditional synthetic control approach by using **interactive fixed effects models** (IFE).

## Key Conceptual Difference from Traditional SC

**Traditional SC**: Creates synthetic control as weighted average of control units
- Weights sum to 1, all positive
- Matches on pre-treatment covariates

**GSynth**: Models outcome as combination of:
- Additive fixed effects (unit + time)
- **Interactive fixed effects**: factors (time-varying) × loadings (unit-specific)


## Preparing Data for GSynth

The `fect` package requires data in long format with unit and time identifiers:

```{r gsynth-prep}
# Prepare data for fect/gsynth

basque_gsynth <- basque %>%
  filter(regionno != 1) %>%  # drop Spain aggregate
  mutate(
    region_id = regionno,
    treated   = as.integer(regionno == 17 & year >= 1970)  # 0 before 1970, 1 from 1970 on
  ) %>%
  select(
    region_id, regionname, year, gdpcap, treated, 
    invest, school.illit, school.prim, school.med, school.high, 
    school.post.high,sec.agriculture, sec.energy, sec.construction, 
    sec.industry, sec.services.venta, sec.services.nonventa,popdens
  ) %>%
  arrange(region_id, year)

# Check data structure
glimpse(basque_gsynth)

panelview(gdpcap ~ treated, data = basque_gsynth,  index = c("region_id","year"), pre.post = TRUE) 

```


## Estimation with Cross-Validation

GSynth automatically selects the number of factors using cross-validation:

```{r gsynth-estimate, cache=TRUE}

gsynth_out <- fect(
  gdpcap ~ treated + invest,
  data = basque_gsynth,
  index = c("region_id", "year"),
  method = "gsynth",  #generalized synthetic control
  force = "two-way",  # Include unit and time fixed effects
  CV = TRUE,          # Cross-validate number of factors
  r = c(0, 5),        # Search 0-5 factors
  se = TRUE,          # Compute standard errors
  vartype = "parametric",  # Parametric bootstrap (good for small Ntr)
  nboots = 500,       # Number of bootstrap replications
  parallel = TRUE,    # Use parallel computing
  seed = 12345
)



# Print summary
gsynth_out$est.att
gsynth_out$est.avg #ATT averaged over all periods
gsynth_out$est.beta #coefficients of the time-varying covariates


#Weights
dim(gsynth_out$wgt.implied)
sort(gsynth_out$wgt.implied)

```


## Visualizing Results

### Treatment Effect Over Time

```{r gsynth-gap, fig.width=10, fig.height=6}
# Gap plot: Treatment effect by period
plot(gsynth_out, 
     type = "gap",
     main = "GSynth: Estimated Treatment Effect on GDP",
     xlab = "Years Relative to Treatment",
     ylab = "Effect on GDP per Capita",
     xlim = c(-10, 27))
```

### Counterfactual Trajectories

```{r gsynth-counterfactual, fig.width=10, fig.height=6}
# Show observed vs counterfactual
plot(gsynth_out,
     type = "counterfactual",
     main = "Observed vs Counterfactual GDP (GSynth)",
     xlab = "Year",
     ylab = "GDP per Capita")
```

### Individual Treatment Effects

```{r gsynth-box, fig.width=10, fig.height=5}
# Box plot of individual effects
plot(gsynth_out,
     type = "box",
     main = "Distribution of Treatment Effects Over Time",
     xlab = "Years Relative to Treatment")
```

### Estimated Factors and Loadings

The interactive fixed effects are the key innovation of GSynth. Let's visualize them:

```{r gsynth-factors, fig.width=10, fig.height=5}
# Plot estimated factors (time-varying components)
plot(gsynth_out,
     type = "factors",
     main = "Estimated Time-Varying Factors",
     xlab = "Year")
```

```{r gsynth-loadings, fig.width=10, fig.height=7}
# Plot factor loadings (unit-specific components)
plot(gsynth_out,
     type = "loadings",
     main = "Estimated Factor Loadings by Region")
```


### Calendar Time Effects

```{r gsynth-calendar, fig.width=10, fig.height=5}
# How effects evolve over calendar time
plot(gsynth_out,
     type = "calendar",
     main = "Treatment Effect by Calendar Year",
     ylim = c(-2, 0))
```


## Placebo Test with GSynth

```{r gsynth-placebo, fig.width=10, fig.height=6}
# Placebo test: pretend treatment happened in 1965
basque_placebo_gsynth <- basque_gsynth %>%
  mutate(
    treated_placebo = ifelse(region_id == 17 & year >= 1965, 1, 0)
  )

gsynth_placebo <- fect(
  gdpcap ~ treated_placebo,
  data = basque_placebo_gsynth %>% filter(year <= 1969),  # Only pre-actual-treatment
  index = c("region_id", "year"),
  method = "gsynth",
  force = "two-way",
  CV = TRUE,
  r = c(0, 5),
  se = TRUE,
  vartype = "parametric",
  nboots = 200,
  parallel = TRUE,
  seed = 12345
)

plot(gsynth_placebo,
     type = "gap",
     main = "Placebo Test: Fake Treatment in 1965",
     xlab = "Years Relative to Fake Treatment")

```

**Interpretation**: No systematic effect before actual treatment, supporting validity.



# Part 3: Augmented Synthetic Control

The **Augmented Synthetic Control** method (Ben-Michael, Feller, and Rothstein 2021) adds a **bias correction term** using an outcome model, making it more robust to misspecification.

## Why Augmented Synth?

Traditional SC can be biased if:
1. No weighted combination of controls perfectly matches the treated unit
2. Important confounders are omitted

**Solution**: Use an outcome regression model to correct for bias, while still leveraging SC's data-driven approach.

```{r augsynth}
# Prepare data for augsynth
basque_aug <- basque %>%
  filter(regionno != 1) %>%  # Remove Spain aggregate
  mutate(
    treated = ifelse(regionno == 17, 1, 0),
    time = year
  )

# Run augmented synthetic control
# Using ridge regression as the outcome model
aug_syn <- augsynth(
  gdpcap ~ treated,
  unit = regionno,
  time = year,
  t_int = 1970,
  data = basque_aug,
  progfunc = "Ridge",  # Outcome model
  scm = TRUE           # Include SC component
)

# Summary
summary(aug_syn)
```

```{r augsynth-plot, fig.width=10, fig.height=7}
# Plot the results
plot(aug_syn)
```



## Understanding the Bias Correction

```{r bias-correction, fig.width=10, fig.height=6}
# The augsynth package provides the bias correction term
# Plot decomposition

att_vec <- summary(aug_syn)$att[,2]
aug_syn_att <- mean(att_vec, na.rm = TRUE)

aug_df <- basque_aug %>%
  filter(regionno == 17) %>%
  select(year, gdpcap) %>%
  arrange(year)

# Create visualization showing SC + bias correction
ggplot() +
  geom_line(data = gaps_df, aes(x = year, y = gap), 
            color = "#0072B2", linewidth = 1.2,
            linetype = "dashed") +
  annotate("text", x = 1985, y = -2, 
           label = "Traditional SC gap", color = "#0072B2", size = 4) +
  geom_vline(xintercept = 1970, linetype = "dashed", color = "red") +
  geom_hline(yintercept = 0, alpha = 0.3) +
  geom_hline(yintercept = aug_syn_att, color = "#D55E00", linewidth = 1.2) +
  annotate("text", x = 1985, y = aug_syn_att + 1, 
           label = "Augmented SC average\n(with bias correction)", 
           color = "#D55E00", size = 4) +
  labs(
    title = "How Augmented Synthetic Control Corrects Bias",
    subtitle = "Augmented SC adds an outcome model to correct for imperfect pre-treatment fit",
    x = "Year",
    y = "Treatment Effect Estimate",
    caption = "The difference reflects the bias correction from the ridge regression outcome model"
  ) +
  theme_minimal()
```





## References

Abadie, Alberto, and Javier Gardeazabal. "The economic costs of conflict: A case study of the Basque Country." *American Economic Review* 93.1 (2003): 113-132.

Abadie, Alberto, Alexis Diamond, and Jens Hainmueller. "Synthetic control methods for comparative case studies: Estimating the effect of California's tobacco control program." *Journal of the American Statistical Association* 105.490 (2010): 493-505.

Arkhangelsky, Dmitry, Susan Athey, David A. Hirshberg, Guido W. Imbens, and Stefan Wager. "Synthetic difference-in-differences." *American Economic Review* 111.12 (2021): 4088-4118.

Ben-Michael, Eli, Avi Feller, and Jesse Rothstein. "The augmented synthetic control method." *Journal of the American Statistical Association* 116.536 (2021): 1789-1803.

Xu, Yiqing. "Generalized synthetic control method: Causal inference with interactive fixed effects models." *Political Analysis* 25.1 (2017): 57-76.

Athey, Susan, Guido Imbens, Zhaonan Qu, and Davide Viviano. "Triply robust panel estimators." arXiv preprint arXiv:2508.21536 (2025).

---

**Session Information**

```{r session-info}
sessionInfo()
```
